Ecological Limits to Forecasting: Estimating Uncertainty in Critical Transitions with Deep Learning

### 2022-05-13 manuscript/manuscript.Rmd
a few refs

<!-- some more verbose bits about bifurcation models probably belongs in discussion instead -->
We will focus the analysis of several different forecasting scenarios based around two classic models in population ecology: 
Robert May's consumer-resource model [@May1979], and the Nicholson-Bailey parasatoid-host model [@Nicholson1935]. 
While population dynamics models used to inform decision making in fields such as fisheries are often much more complex, these simple models nevertheless exhibit rich nonlinear dynamics thought to be typical in ecological systems [@]. 
These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points [] which has had important theoretical and practical management outcomes [@Folke2009; @].
May's model exhibits alternative stable states. 
In this one-dimensional model, transitions between these states can occur due to intrinsic stochasticity, external forcing, or the gradual environmental change that results in a catastrophic saddle-node bifurcation and generates hysteresis. 
The Nicholson-Bailey model is a two species model which contains a Hopf bifurcation, a non-catastrophic bifurcation which either creates or destroys a limit cycle -- a stable oscillatory pattern.

Assessing the accuracy of forecasting methods in the face of such bifurcation dynamics is particularly important question for ecological systems and global environmental change problems.
Bifurcations represent the kind of non-linear responses complex systems can make as the result of slowly changing parameters (i.e. the 'environment' in which those dynamics take place.)
This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analogue in the historical data to date.

No-analog scenarios [@Williams]

@@

If the environment slowly alters one of the parameters (say, the encounter efficiency, `h_t`, in our formulation), one of the stable nodes moves closer and closer to the unstable saddle point, leading to a bifurcation that destroys the stable state, leaving the system to suddenly transition to the alternative stable state.
Saddle-node bifurcations (also known as fold bifurcations) also create a phenomenon known as hysteresis, where it is not sufficient to restore the environment to the previous parameter values to recover the previous state. 
Unlike the Hopf bifurcation which exhibits a continuous transition from a stable node to a small limit cycle that then grows, the saddle-node transition is a discontinuous or so-called 'catastrophic' bifurcation.
Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @Scheffer2001]. 
Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dei2011; @Dei2012] to whole-ecosystem experiments [@Carpenter2011], and postulated as a model for global change [@Barnosky2014].

@@

Perhaps the most difficult of all events to predict are those in which large transitions are predominately driven by a random component. 
An example of such a transition event is possible to observe in May's consumer-resource model, in which a stochastic term occasionally results in a transition between alternative stable states.
In such cases, no forecast can precisely predict when a transition will occur, but it is nonetheless possible to deduce the correct distribution of waiting times knowing the correct model. 
In the case of small noise, transitions are Poisson distributed, such that the distribution of waiting times is roughly exponential [e.g. @vanKampen], though post-hoc the trajectories of such transitions can be mistaken for saddle-node transitions [@prosectorsfallacy].
To consider such cases, we will again use May's alternative stable state model, though this time leaving all parameters fixed.

In this context, predicting the probability of a transition in the future based solely on observations prior to a transition occurring is essentially impossible without additional information constraining the model estimate, as such data is equally consistent with infinitely many models or parameter choices which share the same local linearization about the stable point.  
Unlike the saddle-node bifurcation, there is no slowly warping potential basin which can be detected to inform estimates. 
Thus, in this scenario, rather than considering the problem of predicting the future evolution of a single time series based only on it's historical values, we consider an alternative framing of the task:
we imagine our forecaster has access to historical data from one or more comparable systems which includes a previous stochastic transition event.
Based on this data, our forecaster seeks to identify the distribution of expected transition times for analagous systems starting from the same initial condition. 
This parallels actual practice in which researchers would draw on previous examples of stochastic transitions in a system - lake-ecosystem shifts, disease emergence, changing fire regimes, [].
(Note that such stochastic transitions between alternative stable states can also create oscillatory-like dynamics when stochasticity is sufficiently high enough to drive repeated transitions from one attractor to the other and back again.  In such cases, it might be reasonable to estimate a strictly forward-looking forecast of a single system, predicting the distribution of these transitions.)

>>>>

<!-- some more verbose bits about bifurcation models probably belongs in discussion instead -->
We will focus the analysis of several different forecasting scenarios based around two classic models in population ecology: 
Robert May's consumer-resource model [@May1977], and the Nicholson-Bailey parasatoid-host model [@Nicholson1935]. 
Though these models may appear simple when measured against high-dimensional and parameter rich models found in some management contexts such as fisheries, they can exhibit rich nonlinear dynamics and provide greater capacity to generalize [@Levins1966; @Getz2018]. 
These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points which has had important theoretical and practical management outcomes [@Folke2004; @Fischer2009; @Polasky2011].
May's model exhibits alternative stable states. 
In this one-dimensional model, transitions between these states can occur due to intrinsic stochasticity, external forcing, or the gradual environmental change that results in a catastrophic saddle-node bifurcation and generates hysteresis. 
The Nicholson-Bailey model is a two species model which contains a Hopf bifurcation, a non-catastrophic bifurcation which either creates or destroys a limit cycle -- a stable oscillatory pattern.

Assessing the accuracy of forecasting methods in the face of such bifurcation dynamics is particularly important question for ecological systems and global environmental change problems.
Bifurcations represent the kind of non-linear responses complex systems can make as the result of slowly changing parameters (i.e. the 'environment' in which those dynamics take place.)
This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analog in the historical data to date. 
Forecast skill under such no-analog conditions may be particularly relevant to ecological forecasting in the context of global change [@Williams2007].

@@

If the environment slowly alters one of the parameters (say, the encounter efficiency, `h_t`, in our formulation), one of the stable nodes moves closer and closer to the unstable saddle point, leading to a bifurcation that destroys the stable state, leaving the system to suddenly transition to the alternative stable state.
Saddle-node bifurcations (also known as fold bifurcations) also create a phenomenon known as hysteresis, where it is not sufficient to restore the environment to the previous parameter values to recover the previous state. 
Unlike the Hopf bifurcation which exhibits a continuous transition from a stable node to a small limit cycle that then grows, the saddle-node transition is a discontinuous or so-called 'catastrophic' bifurcation.
Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @Scheffer2001; @Folke2004]. 
Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dai2013; @Dai2015] to whole-ecosystem experiments [@Carpenter2011], and postulated as a model for global change [@Barnosky2012].

@@

Perhaps the most difficult of all events to predict are those in which large transitions are predominately driven by a random component. 
An example of such a transition event is possible to observe in May's consumer-resource model, in which a stochastic term occasionally results in a transition between alternative stable states.
In such cases, no forecast can precisely predict when a transition will occur, but it is nonetheless possible to deduce the correct distribution of waiting times knowing the correct model. 
In the case of small noise, transitions are Poisson distributed, such that the distribution of waiting times is roughly exponential [e.g. @vankampen], though post-hoc the trajectories of such transitions can be mistaken for saddle-node transitions [@prosectorfallacy].
To consider such cases, we will again use May's alternative stable state model, though this time leaving all parameters fixed.

In this context, predicting the probability of a transition in the future based solely on observations prior to a transition occurring is essentially impossible without additional information constraining the model estimate, as such data is equally consistent with infinitely many models or parameter choices which share the same local linearization about the stable point.  
Unlike the saddle-node bifurcation, there is no slowly warping potential basin which can be detected to inform estimates. 
Thus, in this scenario, rather than considering the problem of predicting the future evolution of a single time series based only on it's historical values, we consider an alternative framing of the task:
we imagine our forecaster has access to historical data from one or more comparable systems which includes a previous stochastic transition event.
Based on this data, our forecaster seeks to identify the distribution of expected transition times for analagous systems starting from the same initial condition. 
This parallels actual practice in which researchers would draw on previous examples of stochastic transitions in a system - lake-ecosystem shifts, disease emergence, changing fire regimes, [@Scheffer2001; @Folke2004].
(Note that such stochastic transitions between alternative stable states can also create oscillatory-like dynamics when stochasticity is sufficiently high enough to drive repeated transitions from one attractor to the other and back again.  In such cases, it might be reasonable to estimate a strictly forward-looking forecast of a single system, predicting the distribution of these transitions.)
